Glacial flows, simulated faster

### 2023-05-28 talk-iu/slides.tex
serious talk start

\title[Making ice sheet models scale properly]{Making ice sheet models \\ scale properly}

@@

\date[]{February 2023}

@@

{
  \usebackgroundtemplate{\includegraphics[width=\paperwidth]{../images/gray-british-clark2022.png}}
  \begin{frame}
    \titlepage
  \end{frame}
}

@@

\section{what is an ice sheet model?}

@@

\bigskip
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\includegraphics<1>[height=0.69\textheight]{../images/ant-pittard2021.png}
\only<1>{\par {\scriptsize Antarctic ice sheet}} % (Pittard et al 2021)
\only<2>{\vspace{17mm}}
\includegraphics<2>[width=\textwidth]{../images/ant-schoofhewitt2013.png}
\only<2>{\bigskip \hfill {\tiny (Schoof \& Hewitt 2013)}}
\only<2>{\vspace{4mm}}
\only<2>{\begin{center} {\footnotesize note vertical exaggeration, smooth surface, and rough bed} \end{center} }
\includegraphics<3>[height=0.69\textheight]{../images/alps-seguinot2018.png}
\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum} \hfill {\tiny (Seguinot et al 2018)}}
\includegraphics<4>[height=0.69\textheight]{../images/british-clark2022.png}
\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum} \hfill {\tiny  (Clark et al 2022)}}
\includegraphics<5>[height=0.69\textheight]{../images/not-sea-ice.png}
\only<5>{\par {\footnotesize an ice sheet is \emph{not} sea ice!}}
\end{center}
\end{minipage}
\end{frame}

\begin{frame}{basic facts about glaciers}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{very viscous, incompressible, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon
    \end{itemize}
\item glaciers lie on \emph{topography}
    \begin{itemize}
    \item[$\circ$] sometimes they float  (\emph{ice shelf})
    \end{itemize}
\item glacier geometry and velocity evolve \emph{in contact with climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] sub-shelf melt (when floating)
    \item[$\circ$] calving (into ocean)
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{../images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{frame}{what is an ice sheet model?}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
an \alert{ice sheet model} is a map which simulates an ice sheet in a climate
\end{definition} 
\item at least two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
\text{precipitation minus} \\
\text{melt \& runoff}
\end{pmatrix}$$

@@

\begin{frame}{basic ice sheet model: notation}

\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
\item data $a(t,x,y)$, $b(x,y)$ are defined on a \alert{fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2-> solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] also a fixed domain,
    \item[$\circ$] but $s=b$ where there is no ice
    \end{itemize}
\item<3> $s(t,x,y)$ determines the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$, on which the solution velocity $\bu(t,x,y,z)$ is defined:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$

@@

\begin{frame}{basic ice sheet model: conservation}

\begin{itemize}
\item ice sheet evolution should conserve physical quantities:
    \begin{itemize}
    \item[$\circ$] mass
    \item[$\circ$] momentum
    \item[$\circ$] \st{energy} \hfill $\leftarrow$ \emph{ignored for simplicity in this talk}
    \end{itemize}

\medskip
\item conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] in the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
    \intertext{\item[$\circ$] and on the ice surfaces:}
\text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
\end{align*}

@@

\begin{frame}{basic ice sheet model: strong form = NCP coupled to Stokes}

@@

\begin{frame}{basic ice sheet model: is a DAE system}

@@

\begin{frame}{basic ice sheet model: current research and thinking}

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
    \begin{itemize}
    \item[$\circ$] when pressed, most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model
    \item[$\circ$] well-posedness of the lubrication approximation of the model has been considered; existence proved in (Jouvet \& Bueler 2012)
    \end{itemize}

>>>>

\title{Glacial flows, simulated faster}

@@

\date[]{June 2023}

@@

{
  \usebackgroundtemplate{\includegraphics[height=\paperheight]{../images/fade-moraines-indiana.png}}
  \begin{frame}
    \titlepage
  \end{frame}
}

@@

\section{the basics of glaciers and ice sheets}

\newcommand{\basicglaciers}[1]{
\begin{frame}{basic facts about glaciers #1}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is a \emph{very viscous, incompressible, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more soon \dots
    \end{itemize}
\item glaciers lie on \emph{topography}
    \begin{itemize}
    \item[$\circ$] except sometimes they float on water (floating tongue or ice shelf)
    \end{itemize}
\item a glacier's geometry (\emph{free surface}), and its velocity, \emph{evolve in contact with the climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] sub-shelf melt (when floating)
    \item[$\circ$] calving (into ocean)
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{../images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}
}

\basicglaciers{}

\begin{frame}{glaciers}

\begin{center}
\only<1>{\includegraphics[height=0.8\textheight]{../images/polaris.png}} \only<2>{\includegraphics[height=0.8\textheight]{../images/taku-truffer2016.jpg}}
\only<3>{\vspace{-5mm} \mbox{\includegraphics[height=0.8\textheight]{../images/columbiasentinel2.jpg} \includegraphics[height=0.55\textheight]{../images/natgeocolumbia1910.png}}}
\end{center}

{\scriptsize
\only<1>{Polaris Glacier \hfill {\tiny (Post and LaChappelle 1971)}}
\only<2>{Taku Glacier \hfill {\tiny (M.~Truffer 2016)}}
\only<3>{Columbia Glacier \hfill {\tiny (Sentinel-2B 2018, National Geographic 1910)}}
}
\end{frame}

@@

\bigskip
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\includegraphics<1>[height=0.69\textheight]{../images/ant-pittard2021.png}
\only<1>{\par {\scriptsize Antarctic ice sheet} \hfill {\tiny (Pittard et al 2021)}}
\only<2>{\vspace{17mm}}
\includegraphics<2>[width=\textwidth]{../images/ant-schoofhewitt2013.png}
\only<2>{\vspace{4mm}\par {\scriptsize note smooth surface and rough bed \dots and vertical exaggeration} \hfill {\tiny (Schoof \& Hewitt 2013)}}
\includegraphics<3>[height=0.69\textheight]{../images/alps-seguinot2018.png}
\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum} \hfill {\tiny (Seguinot et al 2018)}}
\includegraphics<4>[height=0.69\textheight]{../images/moraines-indiana.jpg}
\only<4>{\par {\scriptsize moraines in Illinois, Indiana, Ohio} \hfill {\tiny (Larsen 1986 and other sources)}}
\includegraphics<5>[height=0.69\textheight]{../images/laurentide-margold2018.png}
\only<5>{\par {\scriptsize Laurentide ice sheet} \hfill {\tiny (Margold, Stokes, Clark 2018)}}
\end{center}
\end{minipage}
\end{frame}

\begin{frame}{an ice sheet is \emph{not} sea ice!}
\begin{center}
\includegraphics[height=0.8\textheight]{../images/not-sea-ice.png}
\end{center}
\end{frame}

\basicglaciers{\dots again}

@@

\begin{frame}{what is a glacier model?}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
a \alert{glacier} or \alert{ice sheet model} is a \underline{map}

which evolves a glacier in a climate
\end{definition} 
\item at least two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
\text{precipitation minus} \\
\text{melt \& runoff}
\end{pmatrix}$$

@@

\begin{frame}{basic model: notation}

\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
\item data $a(t,x,y)$, $b(x,y)$ are defined on a \alert{fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2-> solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] also a fixed domain,
    \item[$\circ$] but $s=b$ where there is no ice
    \end{itemize}
\item<3-> $s(t,x,y)$ determines the \alert{icy domain} $\Lambda(t) \subset \RR^3$:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$
\item<4> the solution velocity $\bu(t,x,y,z)$ is defined on $\Lambda(t)$

@@

\begin{frame}{basic model: conservation}

\begin{itemize}
\item ice sheet evolution should conserve physical quantities:
    \begin{itemize}
    \item[$\circ$] mass
    \item[$\circ$] momentum
    \item[$\circ$] energy \hfill $\leftarrow$ \emph{ignored for simplicity in this talk}
    \end{itemize}

\medskip
\item conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] in the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t) \\
    \intertext{\item[$\circ$] and on the ice surfaces:}
\text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t) \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t)
\end{align*}

@@

\begin{frame}{basic model: strong form = NCP coupled to Stokes}

@@

\begin{frame}{the basic model is a DAE system}

@@

\begin{frame}{basic model: current research}

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
    \begin{itemize}
    \item[$\circ$] though most researchers would agree NCP-coupled-to-Stokes is indeed the intended model \dots
    \end{itemize}
\item progress has been made on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
    \begin{itemize}
    \item[$\circ$] 1D well-posedness on flat bed (Calvo et al 2003)
    \item[$\circ$] 2D steady-state existence on general beds (Jouvet \& Bueler 2012)
    \item[$\circ$] 2D well-posedness on flat bed (Piersanti \& Temam 2022)
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{basic model: current \emph{thinking}}

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
    \begin{itemize}
    \item[$\circ$] most researchers would agree NCP-coupled-to-Stokes is indeed the intended model!
    \item[$\circ$] progress on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
        \begin{itemize}
        \item 1D well-posedness on flat bed (Calvo et al 2003)
        \item 2D steady-state existence on general beds (Jouvet \& Bueler 2012)
        \item 2D well-posedness on flat bed (Piersanti \& Temam 2022)
        \end{itemize}
    \end{itemize}
